The role of aneuploidy in the evolution of cancer drug resistance

### 2024-05-06 EvolutionaryRescue1.tex
Add files via upload

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to various treatments. 
One evolutionary mechanism by which a population of cancer cells can to adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we examine how aneuploidy impacts the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, and rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells. Additionally, we investigate what effect having a fraction of tumor cells being aneuploid before the onset of therapy have on the probability of evolutionary rescue.  We also approximate the mean recurrence time for the tumor to revert back to its initial size.
We find that aneuploidy can play an important role in the relapse of secondary tumors who have not yet been detected and thus are smaller in size.
% TODO missing something here... ok
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}, caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy. Importantly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to chemotherapeutic agents such as vemurafenib. Reversine-treated cells had higher proliferation rate in the environment altered by anti-cancer drugs compared to wildtype cancer cells.
 Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy, despite having lower fitness before the onset of chemotherapy.
One proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.

An important aspect of aneuploidy is the rate with which cells become aneuploid, which is several orders of magnitude higher than the beneficial mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster when compared to acquiring a mutation, especially when several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer \citep{lee2016effects}.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.

\citet{gunnarsson2020understanding} analyze a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue. Evolutionary rescue in one step in which an initially declining population, after a sudden environmental change, has to acquire a mutation has been studied in the context of population genetics  by \citep{orr2008population,orr2014population}. They analyzed a model where the mutant strain is present in small number at the onset of therapy and concluded that this can significantly enhance the chance that the population will survive.

Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute.  Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We wish to understand what effect does aneuploidy have on the probability of evolutionary rescue when it acts as an intermediary between the wildtype and the mutant cancer cells. We also calculate the mean time that an initially declining tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal} we also take into consideration the effect that standing genetic variation has on the tumor dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to wildtype cells in a drug-free environment.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (left column, see \Cref{figureAneuploidy}):

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (see Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).

@@

We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division. 
\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.21\times10^{-3}$ per chromosome per cell division. \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy which improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed together wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. As a result, the fraction of cancer cells with ``beneficial'' aneuploidy is $f=2\times10^{-3}\times10^{-1}/0.07=0.284\%$ (i.e. $0.284\%$ of pre-treatment cancer cells have the ``beneficial'' aneuploidy).

We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy except trisomy in Chr II and VI as those are the aneuploid cells which are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our analytical calculations, we assume that lineages produced by cells from the initial population divide and die independently of each other.
Of course, the cells will actually compete for resources, but we expect that this can be neglected because the drug will
cause the cell density to rapidly drop far below the carrying capacity at which these interactions are important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity of this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wild type. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the cancer population, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistance for the cancer population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. de-novo genetic variation.}

>>>>

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA}
\affil[*]{Corresponding author: Yoav Ram \textcolor{green}{(e-mail: yoavram@tauex.tau.ac.il)}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue is the process by which a population survives a sudden environmental change that initially causes the population to decline towards extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can have higher fitness in an environment altered by anti-cancer drugs, e.g., because of incomplete pathways targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and moreover, some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we examine how aneuploidy impacts the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the sensitive and resistant cells. Additionally, we investigate the effect on the probability of evolutionary rescue of having a fraction of aneuploid tumor cells before the onset of therapy.  We also approximate the mean recurrence time for the tumor to revert back to its initial size following treatment.
We find that aneuploidy can play an important role in the relapse of smaller secondary tumors who have not yet been detected.
% TODO missing something here... ok
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  Each year approximately 10 million people die from cancer \citep{kocarnik2022cancer}, so understanding the factors that contribute to failure of interventions is of great importance. One hypothesized factor is aneuploidy, where cells are characterized by an imbalanced karyotype and chromosome copy number alterations \citep{schukken2018cin}. Aneuploidy is caused by chromosomal instability, the mitotic process in which cells suffer from chromosome mis-segregation. Importantly, aberrations in chromosome copy number allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective}. Indeed, cancer cells are often aneuploid and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1 and then to chemotherapeutic agents such as vemurafenib. Reversine-treated cells had higher proliferation rate in the environment altered by anti-cancer drugs compared to wildtype cancer cells.
 Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to wildtype cells during chemotherapy, despite having lower fitness before the onset of chemotherapy.
One proposed mechanism through which aneuploidy is able to confer resistance to chemotherapeutics is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}.

An important aspect of aneuploidy is the rate with which cells become aneuploid, which is several orders of magnitude higher than the beneficial mutation rate \citep{bakker2023predicting}. Consequently, a cell exposed to a stress such as chemotherapeutic drugs can acquire aneuploidy faster compared to acquiring a mutation. Morover several proposed anti-cancer drugs elevate the rate of mis-segregation in order to fight cancer cells~\citep{lee2016effects}.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.

\citet{gunnarsson2020understanding} analyzed a model where a tumor consisting of two populations of cancer cells, one drug resistant and the other drug sensitive, is able to evade extinction by cells switching between the two phenotypes through epigenetic mutations. They found that even when the drug resistant type is barely viable the epimutations have the effect of guaranteeing evolutionary rescue. Evolutionary rescue in one step where an initially declining population, after a sudden environmental change, has to acquire a mutation in order to survive extinction has been studied by \citet{orr2008population,orr2014population}. They analyzed a model where the mutant strain is present in small numbers at the onset of therapy and concluded that this can significantly enhance the probability that the population will survive.

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute \textcolor{green}{towards evolutionary rescue}.  Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated using diffusion approximation by \citet{martin2013probability} \textcolor{green}{and tested the predictions of their model with data from experiments with yeast and bacterium}.

Here we build on previous work on evolutionary rescue after a sudden environment change caused by the initiation of chemotherapy. We \textcolor{green}{estimate the} effect of aneuploidy on the probability of evolutionary rescue when it acts as an \textcolor{green}{evolutionary stepping stone} between the \textcolor{green}{sensitive} and the \textcolor{green}{resistant} cancer cells. We also estimate the mean time \textcolor{green}{until the} tumor cell population reaches its pre-treatment size. Given that aneuploidy is present in many tumors even before the onset of therapy \citep{lukow2021chromosomal}, we also consider the effect of \textcolor{green}{pre-treatment} standing genetic variation on the \textcolor{green}{evolutionary} dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the effect that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

\paragraph{Evolutionary model.}
We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible or sensitive and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$.
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing, stationary, or growing or dying only very slowly, that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$, in a sense that we will make precise below. 
We assume that both chromosomal missegregation and mutation occur during the process of mitosis. 
Wildtype cells may divide and then missegregate to become aneuploids at rate $u\lambda_w$. Both aneuploid and wildtype cells may divide and mutate to become mutants at rates $v\lambda_{a}$ and $v\lambda_{w}$, respectively.
To model standing genetic variation, we assume that before the onset of therapy, wildtype cells become aneuploid with rate $\tilde{u}\lambda_w$ (which may differ from $u \lambda_w$) and that aneuploidy confers a fitness cost $s$ in the drug-free environment, that is, we assume that aneuploid cells have an increased death rate compared to wildtype cells in a drug-free environment.
See \Cref{figureAneuploidy} for a schematic representation of the model and \Cref{sampleTrajectories} for sample trajectories of the different genotypes.

@@

The \textcolor{green}{cell population} at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the \textcolor{green}{population} (left column, see \Cref{figureAneuploidy}):

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\left(1-u-v\right)\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(0,+1,0)&:\quad u\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ww_t\quad\left(\text{wildtype cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper we assume that the division rates for wildtype and aneuploid cells can be written as $\lambda_ww_t\left(1-u-v\right)\approx \lambda_ww_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (Table \ref{table1}).
Each iteration of the simulation loop starts with computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction  (see Appendix A).

@@

We assume the mutation rate is $10^{-7}$ per gene per cell division \citep{loeb2001mutator} and since we assume that a single target gene confers resistance to the drug, we use $v=10^{-7}$ per cell division. 
\citet{bakker2023predicting} determined that the missegregation rate must be between $10^{-3}-10^{-2}$ per chromosome per cell division with the optimal value being $6.2\times10^{-3}$ per chromosome per cell division \textcolor{green}{(for $u\ll10^{-3}$ no adaptation happens and for $u\gg10^{-2}$ karyotypes will acquire lethal nullisomies)}. \citet{ippolito2021gene} observed that trisomy in Chr II and VI are most likely to confer increased resistance against the chemotherapeutic agent vemurafenib for A375 cells. We assume that if a tumor is aneuploid then it most likely has trisomy \citep{gisselsson2010generation} in the pre-treatment environment and that cells with more than one trisomy are very unlikely to survive. Additionally, we assume that all trisomies are equally likely and, as a result, we select $u=10^{-2}$ per cell division. For the missegregation rate in the drug-free environment, $\tilde{u}$, we use the lower-end value of $\tilde{u}=2\times10^{-3}$ per cell division, as some drugs increase the rate of aneuploidy~\citep{wang2019molecular,mason2017functional}.

The fitness cost $s$ of aneuploidy before the onset of therapy is difficult to estimate as we are interested in a specific type of aneuploidy that improves the fitness of cancer cells in an environment altered by drugs.
We derive $s$ by using the formula $s=\tilde{u}\lambda_w / f$, where $f$ is the fraction of aneuploid cancer cells. To estimate $f$, we note that \citet{lukow2021chromosomal} mixed wildtype and aneuploid A375 melanoma cells at $50:50$ ratio, cultured them in drug-free environment and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. We obtain the fitness cost $s$ using the formula $s=\abs{\log\left[0.15/(1-0.15)\right]/24}\approx0.07$ per day \citep{chevin2011measuring}. As a result, the fraction of cancer cells with aneuploidy \textcolor{green}{of interest} is $f=2\times10^{-3}\times10^{-1}/0.07=0.28\%$ (i.e., $0.28\%$ of pre-treatment cancer cells have the aneuploidy \textcolor{green}{of interest}).

We note that when we refer to wildtype cancer cells we include those cells that have any aneuploidy \textcolor{green}{other than} trisomy in Chr II and VI as those are the aneuploid cells that are hypothesized to have higher fitness in the environment altered by drugs such as vemurafenib.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In our \textcolor{green}{analysis}, we assume that lineages produced by cells from the initial population divide and die independently of each other.
\textcolor{green}{However}, the cells will compete for resources, but we expect that this \textcolor{green}{competition} can be neglected because the drug will cause the cell density to rapidly drop below the carrying capacity at which \textcolor{green}{competition is most} important.
To test this, we simulate a logistic growth model, with division and death rates given by:
\begin{align*}
\lambda_w' &= \lambda_w , \\
\mu_w' &= \mu_w ,\\
\lambda_a' &= \lambda_a ,\\ 
\mu_a' &= \mu_a + \lambda_a\frac{w+a+m}{K} ,\\
\lambda_m' &= \lambda_m ,\\ 
\mu_m' &= \mu_m + \lambda_m\frac{w+a+m}{K} ,
\end{align*}
where $K$ is the tumor carrying capacity. 
The effective carrying capacity \textcolor{green}{in} this model is $K_e=K\Delta_a/\lambda_a\approx10^6$ for $K=10^8, \lambda_a=0.0901,\mu_a=0.09$, where we define the effective carrying capacity to be the population size at which the aneuploid division rate is equal to the aneuploid death rate.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can then be approximated by one of the following cases, depending on $\Delta_a T^*$, the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_w}}{u\lambda_w} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
These approximations perform very well when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}).
The first line describes the case in which aneuploid cells are still effectively killed by the treatment, but not as quickly as the wildtype. 
In the second case, aneuploid cells are sufficiently resistant that the expected size of each aneuploid lineage is roughly 1.
In both of these cases, aneuploidy increases the probability of rescue by slowing or halting the decrease of the \textcolor{green}{tumor} population \textcolor{green}{size}, allowing more opportunities for producing resistant mutants. 
In the third case, aneuploid cells are sufficiently resistant for the population to re-grow the tumor even without additional resistance mutations.
Notably, in this case the mutant parameters ($v$, $\lambda_m$, and $\Delta_m$) do not affect $N_a^*$ beyond their effect on $T^*$.
In all cases, $N_a^*$ is proportional to $1/u$ such that increasing the missegregation rate $u$ will decrease the threshold tumor size (\Cref{rescue_threshold}B).
Furthermore, increasing the aneuploid growth rate $\Delta_a$ (which appears both in the terms and in the conditions), also reduces the threshold tumor size, with a sharp decrease around $\Delta_a=0$, but the effect is minor when $|\Delta_a|$ is small compared to $T^*$ as this would result in the second case where  $dN_a^*/d\Delta_a=0$ (\Cref{rescue_threshold}A). % TODO these are pretty straightforward - can you glean something more interesting from eq 4? I am still looking for more interesting results from eq4

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. \emph{de-novo} genetic variation.}

In the above we assumed that at the onset of drug treatment, the initial tumor consisted entirely of wildtype cells that are drug sensitive.
However, aneuploidy is likely produced even before \textcolor{green}{the} onset of treatment at some rate $\tilde{u}$ and confers a deleterious fitness effect $s$ in the absence of the drug \citep{replogle2020aneuploidy,giam2015aneuploidy}. Furthermore, the aneuploidy rate in the presence of drugs is likely higher than in their absence, $\tilde{u} < u$ \citep{wang2019molecular,mason2017functional}.
But if the number of cells in the tumor $N$ is large (as expected if the tumor is treated with a drug), there may already be a fraction $f \approx \tilde{u}\lambda_w/s$ of aneuploid cells in the population (here we assume that the drug affects the wildtype death rate but not the division rate and therefore \textcolor{green}{we} use $\lambda_w$ for the wildtype division rate in the drug-free environment).
